---
title: "NYPD Shooting Data"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2025-06-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(caret)
library(ggplot2)
library(reshape2)

```

```{r, message=FALSE, warning=FALSE}
df <- read_csv("NYPD_Shooting_Incident_Data__Historic_.csv")
df <- df %>%
  mutate(
    OCCUR_DATETIME = mdy_hms(paste(OCCUR_DATE, OCCUR_TIME)),
    YEAR = year(OCCUR_DATETIME),
    X_COORD_CD = as.numeric(X_COORD_CD),
    Y_COORD_CD = as.numeric(Y_COORD_CD)
  )
```

##Visualization: Shooting Incidents Over Time by Borough

```{r}
borough_yearly <- df %>%
  filter(!is.na(BORO)) %>%
  group_by(YEAR, BORO) %>%
  summarise(Incidents = n(), .groups = "drop")

ggplot(borough_yearly, aes(x = YEAR, y = Incidents, color = BORO)) +
  geom_line(size = 1.1) +
  geom_point() +
  labs(title = "Shooting Incidents Over Time by Borough", x = "Year", y = "Number of Incidents") +
  theme_minimal()
```

##Visualization: Shootings by Hour of Day

```{r, fig.width=8, fig.height=5}
df %>%
  mutate(HOUR = hour(OCCUR_DATETIME)) %>%
  count(HOUR) %>%
  ggplot(aes(x = HOUR, y = n)) +
  geom_col(fill = "#2c7fb8") +
  labs(title = "Shooting Incidents by Hour of Day",
       x = "Hour of Day (24h format)", y = "Number of Incidents") +
  theme_minimal() +
  scale_x_continuous(breaks = 0:23)
```

## Modeling: Predicting Statistical Murders

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='asis'}
library(caret)

# Filter and prepare data
model_data <- df %>%
  filter(
    !is.na(PERP_AGE_GROUP), PERP_AGE_GROUP != "(null)",
    !is.na(PERP_RACE), PERP_RACE != "(null)",
    !is.na(PERP_SEX), PERP_SEX != "(null)",
    !is.na(BORO), !is.na(LOCATION_DESC),
    !is.na(STATISTICAL_MURDER_FLAG)
  ) %>%
  select(STATISTICAL_MURDER_FLAG, PERP_AGE_GROUP, PERP_SEX, PERP_RACE,
         VIC_AGE_GROUP, VIC_SEX, VIC_RACE, BORO, LOCATION_DESC) %>%
  mutate(across(everything(), as.factor))

# Split into training and test
set.seed(123)
train_index <- createDataPartition(model_data$STATISTICAL_MURDER_FLAG, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data  <- model_data[-train_index, ]

# Train random forest model (safe mtry range)
model <- train(
  STATISTICAL_MURDER_FLAG ~ ., 
  data = train_data,
  method = "rf",
  tuneLength = 3,
  trControl = trainControl(method = "cv", number = 3)
)

# Predict
pred <- predict(model, newdata = test_data)

# Evaluate
cm <- confusionMatrix(pred, test_data$STATISTICAL_MURDER_FLAG)
cat("```\n", paste(capture.output(cm), collapse = "\n"), "\n```")
```

```markdown
## Bias Discussion

**Potential Biases:**

- **Missing Perpetrator Data**: Many entries lack `PERP_RACE`, `PERP_AGE_GROUP`, and `PERP_SEX`, possibly due to unresolved investigations or unreported suspects.

- **Reporting Bias**: Some boroughs have heavier police presence (e.g., Bronx, Brooklyn), which may inflate incident counts compared to under-policed areas.

- **Racial Bias**: Discrepancies between victim and perpetrator race may reflect systemic inequities in crime reporting or enforcement.
